---
title: "Data 607 - Homework 3"
author: "Manolis Manoli"
output:
    pdf_document:
        extra_dependencies: ["geometry", "multicol", "multirow"]
editor_options: 
  chunk_output_type: console
---

#1. Using the 173 majors listed in fivethirtyeight.com’s College Majors dataset [https://fivethirtyeight.com/features/the-economic-guide-to-picking-a-college-major/], provide code that identifies the majors that contain either "DATA" or "STATISTICS"

```{r Stat_data_find,eval=TRUE}

library(httr)
urlRemote  <- "https://raw.githubusercontent.com/"
pathGithub <- "fivethirtyeight/data/master/college-majors/"
fileName   <- "majors-list.csv"


majors <- read.csv(paste0(urlRemote, pathGithub, fileName),header = TRUE)

grep(pattern = 'STATISTICS|DATA', majors$Major, value = TRUE, ignore.case = TRUE)

```

#2 Write code that transforms the data below:

[1] "bell pepper"  "bilberry"     "blackberry"   "blood orange"

[5] "blueberry"    "cantaloupe"   "chili pepper" "cloudberry"  

[9] "elderberry"   "lime"         "lychee"       "mulberry"    

[13] "olive"        "salal berry"

Into a format like this:

c("bell pepper", "bilberry", "blackberry", "blood orange", "blueberry", "cantaloupe", "chili pepper", "cloudberry", "elderberry", "lime", "lychee", "mulberry", "olive", "salal berry")

First we create the start string
```{r startstring,eval=TRUE}

startstr <- '[1] "bell pepper"  "bilberry"     "blackberry"   "blood orange"

[5] "blueberry"    "cantaloupe"   "chili pepper" "cloudberry"  

[9] "elderberry"   "lime"         "lychee"       "mulberry"    

[13] "olive"        "salal berry"'

```
Next we find the positions of all quotation marks
```{r find_quotation,eval=TRUE}
quot = '"'
posi <- gregexpr(patter = quot,text = startstr)
quotpos <- vector()
i <- 1
while (!is.na(posi[[1]][i])){
quotpos[i] <- posi[[1]][i]
i <- i+1
}

```
Then we create a vector of all words
```{r create_word_vec,eval=TRUE}

len <- length(quotpos)/2
wordvect <- vector(length=len)
for (i in 1:len) {
wordvect[i] <- substring(startstr,posi[[1]][2*i-1]+1,posi[[1]][2*i]-1)
i <- i+1
}
print(wordvect)



```
The two exercises below are taken from R for Data Science, 14.3.5.1 in the on-line version:

#3 Describe, in words, what these expressions will match:

    (.)\1\1
    will match characters that are three times in a row
    
    "(.)(.)\\2\\1"
    will match any 4 characters that read the same forward and backward (palindrome)
    
    (..)\1
    two characters repeated 
    
    "(.).\\1.\\1"
    five characters where the first, third and fifth are the same and the second and fourth can be anything
    
    "(.)(.)(.).*\\3\\2\\1"
    6 or more carachters where the first three charachters are the same as the last three in reverse order (abcxyzcba)

#4 Construct regular expressions to match words that:

    Start and end with the same character.
    "^(.).*\\1$"
    
    Contain a repeated pair of letters (e.g. "church" contains "ch" repeated twice.)
    "(..).*\\1"
    
    Contain one letter repeated in at least three places (e.g. "eleven" contains three "e"s.)
    "(.).*\\1.*\\1"
