---
title: "Data 607 - Assignment 9"
author: "Manolis Manoli"
output:
  html_document:
    df_print: paged
  pdf_document: default
editor_options:
  chunk_output_type: console
---

The task is to choose one of the New York Times APIs, construct an interface in R to read in the JSON data, and transform it into an R DataFrame.

Load needed libraries
```{r load-libraries,warning=F}
library(httr)
library(jsonlite)
library(knitr)
library(dplyr)
library(tidyr)
library(plyr)
```

#### 1) create NYT URL

```{r set-URL,eval=T}
url <- "https://api.nytimes.com/svc/"
query <- "movies/v2/reviews/search.json?opening-date=1989-01-01;1990-01-01&api-key="
key  <- "4AfLOCzNQR72CE1EoSJLpRAIP5xV5y4s"
#combine URL and my key into a URL
JSON_URL <- paste0(url,query,key)
```


#### 2) create dataframe from JSON
  
Next we do the same with the JSON format:
```{r JSON,eval=T}
#create a "JSON-list"
JSON <- fromJSON(JSON_URL)

#make JSON into a table
JSON_table <- JSON$results
#make table into data frame
JSON_df <- as.data.frame(JSON_table)

#have a look at the column names and first few rows
names(JSON_df)
head(JSON_df)

```

#### 3) looping through pages 

I used the looping through multiple pages from: https://cran.r-project.org/web/packages/jsonlite/vignettes/json-paging.html   
My attempt failed and I couldn't find a solution to my rownames issue (see below)
This looping works in the colsole, however it will not let me bind it in a markdown (potentially due to limits on API)
```{r looping-pages,eval=F}
#store all pages in a list first

pages <- list()
for(i in 1:10){
  mydata <- fromJSON(paste0(JSON_URL, "&offset=", i))
  message("Retrieving page ", i)
  pages[[i+1]] <- mydata$results
}

#combine all into one
longmovielist <- rbind_pages(pages)
```

#### 4) my looping attempt 

My loop below would not work since merging the resulting data frames would not work since the rownames has the same counting and there seems to be nothing you can to to change the row numbering

Error in .rowNamesDF<-(x, value = value) :
  duplicate 'row.names' are not allowed
In addition: Warning message:
non-unique values when setting 'row.names': ‘1’, ‘10’, ‘11’, ‘12’, ‘13’, ‘14’, ‘15’, ‘16’, ‘17’, ‘18’, ‘19’, ‘2’, ‘20’, ‘3’, ‘4’, ‘5’, ‘6’, ‘7’, ‘8’, ‘9’

```{r looping--broken,eval=F}
#initilizing the loop parameters:
brk=0
i=1
longmovieDF <- JSON_df

#while will break when the resulting JSON file is empty
while(brk==0){
url <- "https://api.nytimes.com/svc/"
query1 <- "movies/v2/reviews/search.json?opening-date=1989-01-01;1990-01-01&offset="
query2 <- "&api-key="
key  <- "4AfLOCzNQR72CE1EoSJLpRAIP5xV5y4s"

offs=20*i

JSON_URL <- paste0(url,query1,offs,query2,key)
#create a "JSON-list"
JSON <- fromJSON(JSON_URL)
#JSON <- fromJSON(JSON_URL)
JSON_table <- JSON$results
#make table into data frame
JSON_df <- as.data.frame(JSON_table)


rownames1 <- seq(from = offs + 1, to = offs + nrow(JSON_df) , by = 1)
rownames(JSON_df) <- rownames1
#rownames(longmovieDF)

rownames(JSON_df)
rownames(longmovieDF)

longmovieDF <- rbind(longmovieDF,JSON_df)

# 
# names(longmovieDF)
# df1 <- as.matrix(JSON_df)
# df2 <- as.matrix(JSON_df)
# longmovieDF <- rbind(df1,df2)
# longmovieDF <- as.data.frame(longmovieDF)        
                
if(nrow(JSON_df) == 0){
  brk=1}

longmovieDF <- rbind(longmovieDF,JSON_mtrx)
i < i+1
i
}

#after we ran all loops we transform back
longmovieDF <- as.data.frame(longmovieDF)  
```
#### 5) Comparison/Results
  
It is surprisingly easy to extract API from the NYT website (the help page is very good)
swiping through pages was a bit more difficult
  

    

GitHub:
https://github.com/chilleundso/DATA607/blob/master/Assignment7/Data607_Assignment7.Rmd

RPubs:
https://rpubs.com/ManolisM/Data607_Assignment7