---
title: "Data 607 - Final Project"
author: "Patrick Maloney & Manolis Manoli"
output: ioslides_presentation
widescreen: true
smaller: true
---

## Introduction

We have chosen to investigate the coronavirus data set provided by John Hopkins University which themselves source the data from DXY, an online platform run by members of the Chinese nedical community.



   
## Loading required libraries

We load all required libraries:

```{r load_libraries, eval=TRUE, results='hide',message=F, include=T}
library(httr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
library(date)
library(tidyverse)
library(jsonlite)
library(kableExtra)
```

## Downloading Covid-19 Data {.smaller}

We download three seperate CSV (for confirmed cases, recoveries and deaths) from the John Hopkins GitHub account and save them as data frames:

 
```{r data_download,eval=TRUE, include=T}

urlRemote  <- "https://raw.githubusercontent.com/"
pathGithub <- "CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
```

```{r ,eval=TRUE, include=T}

fileConfirmGlobal <- "time_series_covid19_confirmed_global.csv"
fileDeathsGlobal <- "time_series_covid19_deaths_global.csv"
fileRecovGlobal <- "time_series_covid19_recovered_global.csv"
```

```{r ,eval=TRUE, include=T}

confsGlobal <- read.csv(paste0(urlRemote, pathGithub, fileConfirmGlobal),header = TRUE)
deathsGlobal <- read.csv(paste0(urlRemote, pathGithub, fileDeathsGlobal),header = TRUE)
recovsGlobal<- read.csv(paste0(urlRemote, pathGithub, fileRecovGlobal),header = TRUE)

```


## Downloading Country Population {.smaller}
We download data which includes population size, ISO2 and ISO3 codes per country to later be able to merge it with other country indices

```{r data_download_countryISO,eval=TRUE}

urlRemote  <- "https://raw.githubusercontent.com/"
pathGithub <- "chilleundso/DATA607/master/Final_Project/"

countryISOfile <- "UID_ISO_FIPS_LookUp_Table.csv"

countryISOfull <- read.csv(paste0(urlRemote, pathGithub, countryISOfile),header = TRUE)

```



## Data Clean-up {.smaller}

As we can see in the above table the data is provide in a long format. In the below section we transform all 3 data frames to take long format:

```{r create3tables,eval=TRUE}
#Creating a long table using gather
confs <- tidyr::gather(confsGlobal, "Date", "Amount", -Province.State, -Country.Region, -Lat, -Long)
confs <- confs %>%
  select(Country.Region, Date, Amount,) %>%
  group_by(Country.Region, Date) %>%
  summarise(Amount = sum(Amount))
#transforming the date column into a date structure:
confs$Date<-as.Date(confs$Date,format='X%m.%d.%y')
#adding a status label to flag data belongs to confirmed data set
confs$Status <- 'confirmed'


```

We do the same transformation for the data sets with recovery and death information

```{r create3tables2,eval=TRUE,include=F}
#Creating a long table using gather
recovs <- tidyr::gather(recovsGlobal, "Date", "Amount", -Province.State, -Country.Region, -Lat, -Long)
recovs <- recovs %>%
select(Country.Region, Date, Amount,) %>%
group_by(Country.Region, Date) %>%
summarise(Amount = sum(Amount))
#transforming the date column into a date structure:
recovs$Date<-as.Date(recovs$Date,format='X%m.%d.%y')
#adding a status label to flag data belongs to recovered data set
recovs$Status <- 'recovered'

#Creating a long table using gather
deaths <- tidyr::gather(deathsGlobal, "Date", "Amount", -Province.State, -Country.Region, -Lat, -Long)
deaths <- deaths %>%
select(Country.Region, Date, Amount,) %>%
group_by(Country.Region, Date) %>%
summarise(Amount = sum(Amount))
#transforming the date column into a date structure:
deaths$Date<-as.Date(deaths$Date,format='X%m.%d.%y')
#adding a status label to flag data belongs to deaddata set
deaths$Status <- 'dead'

kable(head(confs))

```

## Retrieve latest date for each country and widen  {.smaller}

We retrieve the latest date for each country in each of the 3 data frames
```{r latest_per_country,eval=TRUE, message=F}
#retrieve the latest date for each file
require(data.table)

latestconfs <- as.data.table(confs)
latestconfs <- latestconfs[latestconfs[, .I[Date == max(Date)], by=Country.Region]$V1]

latestrecovs <- as.data.table(recovs)
latestrecovs <- latestrecovs[latestrecovs[, .I[Date == max(Date)], by=Country.Region]$V1]

latestdeaths <- as.data.table(deaths)
latestdeaths <- latestdeaths[latestdeaths[, .I[Date == max(Date)], by=Country.Region]$V1]

```

We stack and then widen the 3 data frames

```{r combine_latest_in_master,eval=TRUE}
#we stack all three data sets underneath each other
latestmaster <- rbind(latestconfs,latestrecovs,latestdeaths)
#we then widen the data set to have three new columns one for each amount of confirmed, recovered and dead cases:
latestmasterwide <- tidyr::spread(latestmaster,Status,Amount)

```


## Resulting Master table {.smaller}

```{r show_master,eval=TRUE}
kable(head(latestmasterwide)) %>%
  kable_styling(bootstrap_options = "striped")
```


## Adding Country information and Corona Metrics {.smaller}

```{r add_ISO_master, echo=TRUE}
master <- merge(latestmasterwide, countryISOfull, by.x = "Country.Region", by.y = "Combined_Key",all.x = TRUE)


```

```{r add_measures}
master$death_per_10kcapita <- master$dead / master$Population *10000
master$death_per_confirmed <- master$dead / master$confirmed
```


```{r showfinal}
kable(head(master))%>%
  kable_styling(bootstrap_options = "striped")
```

## Sourcing Country Indices via API and CSV {.smaller}

Sourcing Country Indices via API from numbeo
```{r, message=F,warning=F}


request <- fromJSON("https://numbeo.com/api/rankings_by_country_historical?api_key=x2mmjp3a6651i3&section=12")
quality_life <- request$`2020`

obesity <- read.csv("https://raw.githubusercontent.com/pmalo46/DATA607-1/master/Final_Project/obesity.csv")
indices <- left_join(quality_life, obesity, by = "country")

```

## Country Indices {.smaller}

```{r}
kable(head(indices)) %>%
  kable_styling(bootstrap_options = "striped") %>%
  scroll_box(width = "100%")

```

<!-- Change name of coutry code column to merge with master -->

```{r, include=F}
names(indices)[names(indices) == "iso3166_country_code"] <- "iso2"
names(indices)
```

## Merge the two data sets into Final data set {.smaller}
```{r}
df_merged <- left_join(indices, master, by = "iso2")
kable(head(df_merged))
```
```{r}
df <- subset(df_merged, select = -c(iso3, Country.Region, Date))
head(df)
```

## Comparing death per confirmed case and per capita {.smaller}
```{r}
m_confrimVcapita <- lm(death_per_confirmed ~ death_per_10kcapita, df)
summary(m_confrimVcapita)
```

With an adjusted R squarred of 64% we can see that these two measures of the response to corona do explain a large proportion of their respective variability. Clearly the death per confirmed case highly depends on the amount of testing that is being done in each of the countries but we do see 


```{r}
ggplot(df, aes(x= death_per_confirmed, y= death_per_10kcapita)) + 
                  geom_point() +
                  geom_text(aes(label=country),hjust=0, vjust=-0.5) + 
                  geom_smooth(method='lm', formula= y  ~ x )
```

We can
## Analysis of death per confirmed cases {.smaller}

```{r}
pair_var <- c(2:5, 7:12, 17, 18)
pairs(df[pair_var])
```

judging by these pair plots, it looks like the most interesting correlations for death rates among confirmed cases might be cpi_index, quality_of_life_index, climate_index, obesity_rate, and possibly pollution_index and healthcare_index

```{r}
m_health <- lm(death_per_confirmed ~ healthcare_index, df)
summary(m_health)

m_pol <- lm(death_per_confirmed ~ pollution_index, df)
summary(m_pol)

m_qol <- lm(death_per_confirmed ~ quality_of_life_index, df)
summary(m_qol)

m_clim <- lm(death_per_confirmed ~ climate_index, df)
summary(m_clim)

m_obes <- lm(death_per_confirmed ~ obesity_rate, df)
summary(m_obes)

m_cpi <- lm(death_per_confirmed ~ cpi_index, df)
summary(m_cpi)
```

It seems like climate index is he only variable that has any statistical significance at all. Let's take a closer look at that.

```{r}
ggplot(df, aes(x=climate_index, y=death_per_confirmed)) + geom_point() + geom_smooth(method = "lm", formula = y~x)

```

This suggests that climate may have an effect on death rates. However, Numbeo's climate index statistic is not an average temperature, but rather a "climate-likeability" index, meaning it is more along the lines of a probability that the average person will enjoy this climate.  Thus, warm but temperate climates are more likely to score higher, whereas more extreme or uncomfortable climates score lower.  So this could either mean that the virus may be less lethal in more extreme climates, or it could be that death rates are higher in areas people are more likely to live.


## Analysis of death per capita

We repeat the analysis above with deaths per capita (per 10'000 inhabitants) and compare the results to the above model fitting 

```{r}
m_health2 <- lm(death_per_10kcapita ~ healthcare_index, df)
summary(m_health2)

m_pol2 <- lm(death_per_10kcapita ~ pollution_index, df)
summary(m_pol2)

m_qol2 <- lm(death_per_10kcapita ~ quality_of_life_index, df)
summary(m_qol2)

m_clim2 <- lm(death_per_10kcapita ~ climate_index, df)
summary(m_clim2)

m_obes2 <- lm(death_per_10kcapita ~ obesity_rate, df)
summary(m_obes2)

m_cpi2 <- lm(death_per_10kcapita ~ cpi_index, df)
summary(m_cpi2)
```


```{r}


ggplot(df, aes(x= healthcare_index, y= death_per_10kcapita)) + 
                  geom_point() +
                  geom_text(aes(label=country),hjust=0, vjust=-0.5) + 
                  geom_smooth(method='lm', formula= y  ~ x )

```


```{r}

dfless1 <- dplyr::filter(df, death_per_10kcapita < 1 )
m_health3 <- lm(death_per_10kcapita ~ healthcare_index, dfless1)
summary(m_health3)

ggplot(dfless1, aes(x= healthcare_index, y= death_per_10kcapita)) + 
                  geom_point() +
                  geom_text(aes(label=country),hjust=0, vjust=-0.5) + 
                  geom_smooth(method='lm', formula= y  ~ x )
```

GitHub:
https://github.com/chilleundso/DATA607/blob/master/Project2/Data607_Assignment5_Corona.Rmd

RPubs:
https://rpubs.com/ManolisM/Project2-1

