---
title: "Data 607 - Final Project"
author: "Patrick Maloney & Manolis Manoli"
output:
  html_document:
    df_print: paged
  pdf_document: default
  chunk_output_type: inline

---


We have chosen to investigate the coronavirus data set provided by John Hopkins University which themselves source the data from DXY, an online platform run by members of the Chinese nedical community.

We combine this data with data sourced through the API of Numbeo which provides country information on various indicators, including health care and quality of life.

Our goal is to investigate wether we can identify any correlation between any of these indicators and either the mortality rate (i.e. death per confirmed case) and the death per capita.
   
#### 1) Loading required libraries and downloading data

We load all required libraries:

```{r load_libraries, eval=TRUE, results='hide',message=F}
library(httr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
library(date)
library(tidyverse)
library(jsonlite)
```

We download three seperate CSV (for confirmed cases, recoveries and deaths) globally from the John Hopkins GitHub account and save them as data frames:


```{r data_download,eval=TRUE}

urlRemote  <- "https://raw.githubusercontent.com/"
pathGithub <- "CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"

fileConfirmGlobal <- "time_series_covid19_confirmed_global.csv"
# fileConfirmUS <- "time_series_covid19_confirmed_US.csv"
fileDeathsGlobal <- "time_series_covid19_deaths_global.csv"
# fileDeathsUS <- "time_series_covid19_deaths_US.csv"
fileRecovGlobal <- "time_series_covid19_recovered_global.csv"


confsGlobal <- read.csv(paste0(urlRemote, pathGithub, fileConfirmGlobal),header = TRUE)
# confsUS <- read.csv(paste0(urlRemote, pathGithub, fileConfirmUS),header = TRUE)
deathsGlobal <- read.csv(paste0(urlRemote, pathGithub, fileDeathsGlobal),header = TRUE)
# deathsUS <- read.csv(paste0(urlRemote, pathGithub, fileDeathsUS),header = TRUE)
recovsGlobal<- read.csv(paste0(urlRemote, pathGithub, fileRecovGlobal),header = TRUE)
```

Download a table that will enable us to merge a unique country iso code and the population so that we can combine the data with the numbeo API data.

```{r data_download_countryISO,eval=TRUE}


urlRemote  <- "https://raw.githubusercontent.com/"
pathGithub <- "chilleundso/DATA607/master/Final_Project/"

countryISOfile <- "UID_ISO_FIPS_LookUp_Table.csv"

countryISOfull <- read.csv(paste0(urlRemote, pathGithub, countryISOfile),header = TRUE)

```



#### 2) Data Clean-up

As we can see in the above table the data is provide in a long format. In the below section we transform all 3 data frames to take long format:

```{r create3tables,eval=TRUE}
#Creating a long table using gather
confs <- tidyr::gather(confsGlobal, "Date", "Amount", -Province.State, -Country.Region, -Lat, -Long)
confs <- confs %>%
  select(Country.Region, Date, Amount,) %>%
  group_by(Country.Region, Date) %>%
  summarise(Amount = sum(Amount))
#transforming the date column into a date structure:
confs$Date<-as.Date(confs$Date,format='X%m.%d.%y')
#adding a status label to flag data belongs to confirmed data set
confs$Status <- 'confirmed'

#Creating a long table using gather
recovs <- tidyr::gather(recovsGlobal, "Date", "Amount", -Province.State, -Country.Region, -Lat, -Long)
recovs <- recovs %>%
select(Country.Region, Date, Amount,) %>%
group_by(Country.Region, Date) %>%
summarise(Amount = sum(Amount))
#transforming the date column into a date structure:
recovs$Date<-as.Date(recovs$Date,format='X%m.%d.%y')
#adding a status label to flag data belongs to recovered data set
recovs$Status <- 'recovered'

#Creating a long table using gather
deaths <- tidyr::gather(deathsGlobal, "Date", "Amount", -Province.State, -Country.Region, -Lat, -Long)
deaths <- deaths %>%
select(Country.Region, Date, Amount,) %>%
group_by(Country.Region, Date) %>%
summarise(Amount = sum(Amount))
#transforming the date column into a date structure:
deaths$Date<-as.Date(deaths$Date,format='X%m.%d.%y')
#adding a status label to flag data belongs to deaddata set
deaths$Status <- 'dead'

kable(head(confs))

```

```{r latest_per_country,eval=TRUE}
#retrieve the latest date for each file
require(data.table)

latestconfs <- as.data.table(confs)
latestconfs <- latestconfs[latestconfs[, .I[Date == max(Date)], by=Country.Region]$V1]

latestrecovs <- as.data.table(recovs)
latestrecovs <- latestrecovs[latestrecovs[, .I[Date == max(Date)], by=Country.Region]$V1]

latestdeaths <- as.data.table(deaths)
latestdeaths <- latestdeaths[latestdeaths[, .I[Date == max(Date)], by=Country.Region]$V1]

```

```{r combine_latest_in_master,eval=TRUE}
#we stack all three data sets underneath each other
latestmaster <- rbind(latestconfs,latestrecovs,latestdeaths)
#we then widen the data set to have three new columns one for each amount of confirmed, recovered and dead cases:
latestmasterwide <- tidyr::spread(latestmaster,Status,Amount)

```


We merge the wide master data set with the country code and population

```{r add_ISO_master, echo=TRUE}
master <- merge(latestmasterwide, countryISOfull, by.x = "Country.Region", by.y = "Combined_Key",all.x = TRUE)

kable(head(master))
```

We add the death per confirmed case and a death per 10'000 inhabitants measure

```{r add_measures}
master$death_per_10kcapita <- master$dead / master$Population *10000
master$death_per_confirmed <- master$dead / master$confirmed
```


Merge with numbeo data
```{r}

request <- fromJSON("https://numbeo.com/api/rankings_by_country_historical?api_key=x2mmjp3a6651i3&section=12")
quality_life <- request$`2020`

obesity <- read.csv("https://raw.githubusercontent.com/pmalo46/DATA607-1/master/Final_Project/obesity.csv")
indices <- left_join(quality_life, obesity, by = "country")

kable(indices)
names(indices)
```
Change name of coutry code column to merge with master
```{r}
names(indices)[names(indices) == "iso3166_country_code"] <- "iso2"
names(indices)
```

Merge the two data sets
```{r}
df_merged <- left_join(indices, master, by = "iso2")
kable(head(df_merged))
```
```{r}
df <- subset(df_merged, select = -c(iso3, Country.Region, Date))
head(df)
```

## Comparing death per confirmed case and per capita
```{r}
m_confrimVcapita <- lm(death_per_confirmed ~ death_per_10kcapita, df)
summary(m_confrimVcapita)
```

With an adjusted R squarred of 64% we can see that these two measures of the response to corona do explain a large proportion of their respective variability. We need to keep in mind that, cleary, the death per confirmed case highly depends on the amount of testing that is being done in each of the countries.


```{r,warning=F}
ggplot(df, aes(x= death_per_confirmed, y= death_per_10kcapita)) + 
                  geom_point() +
                  geom_text(aes(label=country),hjust=0, vjust=-0.5) + 
                  geom_smooth(method='lm', formula= y  ~ x )
```


## Analysis of death per confirmed cases

We start by investigating all scatterplots:

```{r}
pair_var <- c(2:5, 7:12, 17, 18)
pairs(df[pair_var])
```

judging by these pair plots, it looks like the most interesting correlations for death rates among confirmed cases might be cpi_index, quality_of_life_index, climate_index, obesity_rate, and possibly pollution_index and healthcare_index

```{r}
m_health <- lm(death_per_confirmed ~ healthcare_index, df)
summary(m_health)

m_pol <- lm(death_per_confirmed ~ pollution_index, df)
summary(m_pol)

m_qol <- lm(death_per_confirmed ~ quality_of_life_index, df)
summary(m_qol)

m_clim <- lm(death_per_confirmed ~ climate_index, df)
summary(m_clim)

m_obes <- lm(death_per_confirmed ~ obesity_rate, df)
summary(m_obes)

m_cpi <- lm(death_per_confirmed ~ cpi_index, df)
summary(m_cpi)
```

It seems like climate index is he only variable that has any statistical significance at all. Let's take a closer look at that.

```{r,warning=F}
ggplot(df, aes(x=climate_index, y=death_per_confirmed)) + geom_point() + geom_smooth(method = "lm", formula = y~x)

```

This suggests that climate may have an effect on death rates. However, Numbeo's climate index statistic is not an average temperature, but rather a "climate-likeability" index, meaning it is more along the lines of a probability that the average person will enjoy this climate.  Thus, warm but temperate climates are more likely to score higher, whereas more extreme or uncomfortable climates score lower.  So this could either mean that the virus may be less lethal in more extreme climates, or it could be that death rates are higher in areas people are more likely to live.


## Analysis of death per capita

We repeat the analysis above with deaths per capita (per 10'000 inhabitants) and compare the results to the above model fitting 

```{r}
m_health2 <- lm(death_per_10kcapita ~ healthcare_index, df)
summary(m_health2)

m_pol2 <- lm(death_per_10kcapita ~ pollution_index, df)
summary(m_pol2)

m_qol2 <- lm(death_per_10kcapita ~ quality_of_life_index, df)
summary(m_qol2)

m_clim2 <- lm(death_per_10kcapita ~ climate_index, df)
summary(m_clim2)

m_obes2 <- lm(death_per_10kcapita ~ obesity_rate, df)
summary(m_obes2)

m_cpi2 <- lm(death_per_10kcapita ~ cpi_index, df)
summary(m_cpi2)
```


Given that a lot of focus during the crisis has been on the health care system we will investigate this relationship.
  
With an adjusted R squared of 7% we cannot conclude that health care rating is a specifically strong indicator of the amount of people that die in a country. However, we want to investigate this further to see if we can see find a reason for why it could be such a small predictive power.


```{r,warning=F}


ggplot(df, aes(x= healthcare_index, y= death_per_10kcapita)) + 
                  geom_point() +
                  geom_text(aes(label=country),hjust=0, vjust=-0.5) + 
                  geom_smooth(method='lm', formula= y  ~ x ) 

```

We can see that contrary to what we would have anticipated we have a potivie correlation between health care rating and the deaths per capita. The countries that stick out especially are Belgium, Spain, Italy, the UK and France. THey all have a death per capita of more than 4 per 10 thousand. while having a relatively well rated health care system.
It seems like there is a large divide between the European countries and the rest of the world (specifically the Asian countries). Looking at the countries with the best rated healthcare we clearly see that Asian countries have a much lower death rate (less than 5 per million inhabitants) while the Western countries with similar ratings have 20 to 100 times higher rate of death per capita. THis might have more to do with the response and ability to lock down the economy effectively versus western countries:

```{r,warning=F}
head(arrange(df, desc(healthcare_index))[,c(1,4,17)])


``` 

Finally, we need to add that this is not the end of this epidemic and numbers will surely be updated and corrected for the foreseeable future.

