---
title: "Data 607 - Final Project"
author: "Patrick Maloney & Manolis Manoli"
output:
  html_document:
    df_print: paged
  pdf_document: default
editor_options:
  chunk_output_type: console
---


We have chosen to investigate the coronavirus data set provided by John Hopkins University which themselves source the data from DXY, an online platform run by members of the Chinese nedical community.

COVID-19, the scientific name of the coronavirus, had its first recorded case in Wuhan, China in December 2019 and has since across the world with more than 100,000 cases globally as of early March.

We will start by looking at general trends of the spread of the virus, followed by a look at the mortality rate. We finish by investigating the growth rate of confirmed cases, specifically in the most mature case, China, to see if we can find any indication about growth there and make any conclusions about the future of the virus in the rest of the world.

  <br>
  <br>
   
#### 1) Loading required libraries and downloading data

We load all required libraries:

```{r load_libraries, eval=TRUE, results='hide',message=F}
library(httr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
library(date)
library(tidyverse)
```

We download there seperate CSV (for confirmed cases, recoveries and deaths) from the John Hopkins GitHub account and save them as data frames:

```{r data_download,eval=TRUE}

urlRemote  <- "https://raw.githubusercontent.com/"
pathGithub <- "CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"

fileConfirmGlobal <- "time_series_covid19_confirmed_global.csv"
# fileConfirmUS <- "time_series_covid19_confirmed_US.csv"
fileDeathsGlobal <- "time_series_covid19_deaths_global.csv"
# fileDeathsUS <- "time_series_covid19_deaths_US.csv"
fileRecovGlobal <- "time_series_covid19_recovered_global.csv"


confsGlobal <- read.csv(paste0(urlRemote, pathGithub, fileConfirmGlobal),header = TRUE)
# confsUS <- read.csv(paste0(urlRemote, pathGithub, fileConfirmUS),header = TRUE)
deathsGlobal <- read.csv(paste0(urlRemote, pathGithub, fileDeathsGlobal),header = TRUE)
# deathsUS <- read.csv(paste0(urlRemote, pathGithub, fileDeathsUS),header = TRUE)
recovsGlobal<- read.csv(paste0(urlRemote, pathGithub, fileRecovGlobal),header = TRUE)
```


```{r data_download_countryISO,eval=TRUE}

# urlRemote  <- "https://raw.githubusercontent.com/"
# pathGithub <- "CSSEGISandData/COVID-19/master/csse_covid_19_data/"

urlRemote  <- "https://raw.githubusercontent.com/"
pathGithub <- "chilleundso/DATA607/master/Final_Project/"

countryISOfile <- "UID_ISO_FIPS_LookUp_Table.csv"

countryISOfull <- read.csv(paste0(urlRemote, pathGithub, countryISOfile),header = TRUE)
# countryISO <- unique(countryISOfull[,c(3,8)])
# countryISOmapping <- unique(setDT(countryISO)[order(Country_Region,iso3)], by = "Country_Region")
```



#### 2) Data Clean-up

As we can see in the above table the data is provide in a long format. In the below section we transform all 3 data frames to take long format:

```{r create3tables,eval=TRUE}
#Creating a long table using gather
confs <- tidyr::gather(confsGlobal, "Date", "Amount", -Province.State, -Country.Region, -Lat, -Long)
confs <- confs %>%
  select(Country.Region, Date, Amount,) %>%
  group_by(Country.Region, Date) %>%
  summarise(Amount = sum(Amount))
#transforming the date column into a date structure:
confs$Date<-as.Date(confs$Date,format='X%m.%d.%y')
#adding a status label to flag data belongs to confirmed data set
confs$Status <- 'confirmed'

#Creating a long table using gather
recovs <- tidyr::gather(recovsGlobal, "Date", "Amount", -Province.State, -Country.Region, -Lat, -Long)
recovs <- recovs %>%
select(Country.Region, Date, Amount,) %>%
group_by(Country.Region, Date) %>%
summarise(Amount = sum(Amount))
#transforming the date column into a date structure:
recovs$Date<-as.Date(recovs$Date,format='X%m.%d.%y')
#adding a status label to flag data belongs to recovered data set
recovs$Status <- 'recovered'

#Creating a long table using gather
deaths <- tidyr::gather(deathsGlobal, "Date", "Amount", -Province.State, -Country.Region, -Lat, -Long)
deaths <- deaths %>%
select(Country.Region, Date, Amount,) %>%
group_by(Country.Region, Date) %>%
summarise(Amount = sum(Amount))
#transforming the date column into a date structure:
deaths$Date<-as.Date(deaths$Date,format='X%m.%d.%y')
#adding a status label to flag data belongs to deaddata set
deaths$Status <- 'dead'

# kable(head(confs))
```

```{r latest_per_country,eval=TRUE}
#retrieve the latest date for each file
require(data.table)

latestconfs <- as.data.table(confs)
latestconfs <- latestconfs[latestconfs[, .I[Date == max(Date)], by=Country.Region]$V1]

latestrecovs <- as.data.table(recovs)
latestrecovs <- latestrecovs[latestrecovs[, .I[Date == max(Date)], by=Country.Region]$V1]

latestdeaths <- as.data.table(deaths)
latestdeaths <- latestdeaths[latestdeaths[, .I[Date == max(Date)], by=Country.Region]$V1]

```

```{r combine_latest_in_master,eval=TRUE}
#we stack all three data sets underneath each other
latestmaster <- rbind(latestconfs,latestrecovs,latestdeaths)
#we then widen the data set to have three new columns one for each amount of confirmed, recovered and dead cases:
latestmasterwide <- tidyr::spread(latestmaster,Status,Amount)

```



```{r add_ISO_master,eval=TRUE}
master <- merge(latestmasterwide, countryISOfull, by.x = "Country.Region", by.y = "Combined_Key",all.x = TRUE)
```

```{r add_measures,eval=TRUE}
master$death_per_10kcapita <- master$dead / master$Population *10000
master$death_per_conirmed <- master$dead / master$confirmed
```

```{r scatter,eval=TRUE}
ggplot(master, aes(x=death_per_10kcapita, y=death_per_conirmed)) + geom_point()
```

```{r}
library(jsonlite)
library(tidyverse)

request <- fromJSON("https://numbeo.com/api/rankings_by_country_historical?api_key=x2mmjp3a6651i3&section=12")
quality_life <- request$`2020`

obesity <- read.csv("https://raw.githubusercontent.com/pmalo46/DATA607-1/master/Final_Project/obesity.csv")
indices <- left_join(quality_life, obesity, by = "country")
indices
```




GitHub:
https://github.com/chilleundso/DATA607/blob/master/Project2/Data607_Assignment5_Corona.Rmd

RPubs:
https://rpubs.com/ManolisM/Project2-1
